\documentclass[conference]{IEEEtran}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{url}
\usepackage{xspace}

\newcommand{\sys}{GossipRoboFL\xspace}

\begin{document}

\title{GossipRoboFL: Byzantine-Robust Gossip Federated Learning\\for Decentralized Robot Swarms}

\author{
\IEEEauthorblockN{Logan Chu,\; William Zhang,\; Jay Warrier,\; William Vo}
\IEEEauthorblockA{Team GossipRoboFL\quad
CS 390: Distributed Systems, Duke University, Spring 2026}
}

\maketitle

\begin{abstract}
Decentralized robot swarms must learn collaboratively without a central
coordinator.
We built \sys, a fully peer-to-peer federated learning system for 50-robot
swarms operating on dynamic, proximity-based communication graphs.
Robots train locally on heterogeneous image data and share model weights only
with physical neighbors via gossip.
Using the SSClip Byzantine-robust aggregator, \sys sustains 75.7\%
classification accuracy on CIFAR-10 under 20\% malicious robots---within
6~pp of centralized FedAvg with no server---and converges even at spectral
gaps as low as $\lambda_2{=}0.055$.
\end{abstract}

\section{Goals}

Traditional federated learning (FedAvg~\cite{mcmahan2017}) relies on a
central parameter server that is both a communication bottleneck and a single
point of failure.
Robot swarms operating in GPS-denied or infrastructure-free environments
cannot assume persistent server connectivity.
\sys pursues four distributed-systems goals:

\begin{enumerate}
\item \textbf{Serverless coordination.}
Replace the central aggregator with pure peer-to-peer gossip averaging: each
robot pushes its model to $k{=}4$ physical neighbors per round.
\item \textbf{Byzantine fault tolerance.}
Withstand sign-flip adversaries corrupting up to 30\% of the swarm using the
Self-Centred Clipping aggregator (SSClip)~\cite{fraboni2022}.
\item \textbf{Topology sensitivity.}
Characterize how communication range $r$ and algebraic connectivity
$\lambda_2$ (second eigenvalue of the graph Laplacian) govern convergence.
\item \textbf{Data heterogeneity robustness.}
Maintain performance under Dirichlet-partitioned non-IID data
($\alpha \in [0.1, 10.0]$) and straggler clients with variable local epochs.
\end{enumerate}

\section{What We Built}

\subsection{Architecture}

\sys comprises six Python modules under \texttt{src/}:

\textbf{\texttt{simulator.py}} implements \texttt{GossipSimulator}
(decentralized) and \texttt{FedAvgSimulator} (centralized baseline).
Each synchronous round: robots train locally via Ray/\texttt{ThreadPoolExecutor},
each robot gossips to $k$ neighbors, weights are aggregated by the selected
algorithm, and the dynamic topology is rebuilt every 5~rounds.
Metrics (accuracy, $\lambda_2$, cumulative communication bytes) are logged
per round to JSON.

\textbf{\texttt{encounter\_gossip.py}} implements
\texttt{EncounterGossipSimulator} for DTN-style operation: gossip fires only
when robots physically enter communication range, asynchronously with respect
to training.
It tracks graph fragmentation (\texttt{num\_components},
\texttt{mean\_isolation\_steps}), capturing the CAP-theorem tradeoff between
model consistency and continued local availability during network partitions.

\textbf{\texttt{client.py}} models one \texttt{RobotClient} running local SGD
on a Dirichlet-partitioned CIFAR-10 shard.
Byzantine robots inject attacks post-training before broadcasting.

\textbf{\texttt{gossip.py}} provides three aggregators:
\texttt{gossip\_mean} (no resilience), \texttt{clipped\_gossip} (clips
neighbor deltas to an $L_2$ ball of radius $\tau$~\cite{farhadkhani2022}),
and \texttt{ssclip} (reconstructs accepted weights after clipping; drift
$\leq\!\tau$ regardless of Byzantine fraction $f$~\cite{fraboni2022}).

\textbf{\texttt{topology.py}} manages robot positions on a unit square with
random-geometric connectivity.
Robots take Gaussian mobility steps ($\sigma{=}0.05$ per round);
$\lambda_2$ is recomputed via \texttt{numpy.linalg.eigvalsh} on the explicit
Laplacian.

\textbf{\texttt{attacks.py}} provides five attack profiles:
\texttt{sign\_flip}, \texttt{random\_noise}, \texttt{label\_flip},
\texttt{gaussian\_perturb}, and an adaptive anti-centroid
(\texttt{partial\_knowledge}).

The model is \texttt{SmallCNNCifar}: a 3-block Conv--BN--ReLU network
(${\approx}487$k parameters, ${\approx}2$\,MB per message), chosen so that
per-round gossip traffic remains tractable at $N{=}50$, $k{=}4$.

\subsection{Infrastructure}

All experiments ran on a SLURM cluster with NVIDIA GPUs under a managed
\texttt{conda} environment.
Five SLURM job scripts drove the Byzantine, topology, heterogeneity, FedAvg
baseline, and encounter ablations.
Each condition executed 200 gossip rounds with $N{=}50$ robots and logged
per-round metrics to JSON (${\approx}89$\,KB each, 22~files total).
Convergence curves, heatmaps, and per-client accuracy distributions were
generated in \texttt{notebooks/analysis.ipynb}.

\section{Research Questions}

\textbf{Q1:} How do Byzantine fraction $f$ and aggregation algorithm interact
to determine accuracy under a sign-flip attack?

\textbf{Q2:} How does communication range $r$ (and spectral gap $\lambda_2$)
govern gossip convergence under dynamic robot mobility?

\textbf{Q3:} How resilient is gossip FL to Dirichlet data heterogeneity
($\alpha$) and client straggler effects?

\section{Experiments \& Results}

\textbf{Setup:} $N{=}50$ robots, CIFAR-10, Dirichlet $\alpha{=}0.5$,
200~rounds, fanout $k{=}4$, CUDA.
The centralized FedAvg baseline achieves \textbf{81.5\%} accuracy (no
Byzantine robots, no topology constraint).

\subsection{Q1: Byzantine Robustness}

Table~\ref{tab:byz} shows final honest-robot accuracy for three aggregators.
Mean averaging collapses to near-random accuracy (${\approx}10\%$) at
$f{\geq}20\%$---a failure mode that directly mirrors Byzantine consensus
impossibility without sufficient redundancy~\cite{lamport1982}.
ClippedGossip degrades from 75.2\% to 52.3\% at $f{=}10\%$.
SSClip sustains \textbf{75.7\%} at $f{=}20\%$: a \textbf{64.7~pp}
improvement over Mean at the same fault level.
Notably, SSClip under 20\% Byzantine ($75.7\%$) outperforms ClippedGossip
under only 10\% Byzantine ($52.3\%$), demonstrating a qualitatively
stronger fault tolerance guarantee.

\begin{table}[h]
\centering
\caption{Final Honest-Robot Accuracy vs.\ Byzantine Fraction
($N{=}50$, $\alpha{=}0.5$, 200 rounds)}
\label{tab:byz}
\setlength{\tabcolsep}{5pt}
\begin{tabular}{@{}lcccc@{}}
\toprule
Aggregator & $f{=}0\%$ & $f{=}10\%$ & $f{=}20\%$ & $f{=}30\%$ \\
\midrule
Mean          & 76.3\% & 18.6\% & 11.0\% & 10.0\% \\
ClippedGossip & 75.2\% & 52.3\% & ---    & ---    \\
SSClip        & ${\approx}75\%$ & --- & \textbf{75.7\%} & --- \\
\midrule
\multicolumn{5}{l}{\footnotesize{---: condition not run; FedAvg (centralized, no attack): 81.5\%}}\\
\bottomrule
\end{tabular}
\end{table}

\subsection{Q2: Network Topology and Spectral Gap}

Table~\ref{tab:topo} sweeps communication range $r$ under dynamic mobility.
Spectral gap $\lambda_2$ grows 170${\times}$ (from~$0.055$ to~$9.29$) as
$r$ increases from~$0.2$ to~$0.6$, yet final accuracy shifts by only
\textbf{3.4~pp}.
At $r{=}0.2$ ($\lambda_2{=}0.055$), the random-geometric graph is
near-disconnected yet gossip still converges to 74.1\%---demonstrating
\emph{partition tolerance} analogous to eventual consistency.
Accuracy peaks at $r{=}0.5$ ($\lambda_2{=}5.1$) and shows no further
improvement with densification, consistent with the $O(1/\lambda_2)$
mixing-time bound saturating once $\lambda_2 \gg 1$.

\begin{table}[h]
\centering
\caption{Topology Ablation: Comm.\ Range, Spectral Gap, Accuracy
(SSClip, $f{=}20\%$, 200 rounds)}
\label{tab:topo}
\begin{tabular}{@{}cccc@{}}
\toprule
Range $r$ & $\lambda_2$ & Accuracy & Density \\
\midrule
0.2 & 0.055 & 74.1\% & Very sparse \\
0.3 & 0.553 & 76.5\% & Sparse \\
0.4 & 1.894 & 75.9\% & Medium \\
0.5 & 5.079 & \textbf{77.5\%} & Dense \\
0.6 & 9.291 & 76.5\% & Very dense \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Q3: Data Heterogeneity}

Table~\ref{tab:noniid} varies Dirichlet $\alpha$.
The \textbf{15.4~pp} accuracy gap from $\alpha{=}0.1$ to $\alpha{=}10.0$
confirms that data distribution---not topology---is the dominant performance
driver in gossip FL.
At $\alpha{=}0.1$, each robot sees ${\approx}1$--2 of 10 classes; divergent
local gradients overwhelm gossip averaging faster than SSClip can reconcile
them.
Heterogeneous straggler clients (1--5 variable local epochs) perform within
\textbf{0.3~pp} of homogeneous clients, suggesting SSClip's $\tau$-bounded
clipping absorbs staleness-induced gradient variance without requiring
synchronization.

\begin{table}[h]
\centering
\caption{Data Heterogeneity Ablation (SSClip, $f{=}20\%$, $N{=}50$)}
\label{tab:noniid}
\begin{tabular}{@{}lcl@{}}
\toprule
Dirichlet $\alpha$ & Accuracy & Client Type \\
\midrule
0.1 (severe non-IID) & 60.4\% & Homogeneous \\
0.5 (default)        & 75.7\% & Homogeneous \\
1.0                  & 78.1\% & Homogeneous \\
10.0 (near IID)      & 79.8\% & Homogeneous \\
0.5                  & 75.9\% & Heterogeneous (stragglers) \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Communication cost:} At $k{=}4$, $N{=}50$, each round transfers
${\approx}400$\,MB of model weights---linear in $k$, with no server I/O
bottleneck, compared to FedAvg's star topology where server bandwidth scales
with $N$.

\section{Conclusion and Debrief}

\textbf{Key findings.}
SSClip enables Byzantine-robust gossip FL within 6~pp of centralized FedAvg
with zero server infrastructure.
Network topology matters only at the sparse extreme ($\lambda_2 {<} 0.5$);
beyond this threshold, convergence is effectively topology-agnostic.
Data heterogeneity ($\alpha$) is the dominant accuracy driver; straggler
effects are negligible.

\textbf{Distributed systems connections.}
Gossip protocols are a classical DS primitive for epidemic broadcast and
push-sum averaging~\cite{kempe2003}.
\sys demonstrates their application in Byzantine fault-tolerant model
averaging---a direct analog of Byzantine consensus in asynchronous
message-passing networks.
The encounter simulator embodies the CAP tradeoff: partitioned robots
sacrifice consistency (stale model weights) for availability (continued local
training).
The spectral gap $\lambda_2$ serves as the network's algebraic proxy for
consensus round complexity, connecting graph theory directly to convergence
guarantees.

\textbf{Obstacles encountered.}
\begin{enumerate}
\item \textit{Windows/Ray incompatibility.}
Ray's \texttt{fork}-based worker pool is unsupported on Windows~11.
We fell back to \texttt{ThreadPoolExecutor} for local development and ran Ray
only on the Linux SLURM cluster.
\item \textit{SLURM job preemption.}
GPU preemption produced partial result files.
Per-condition output naming conventions allowed us to identify and safely
resubmit incomplete conditions.
\item \textit{Spectral gap on disconnected graphs.}
NetworkX's \texttt{algebraic\_connectivity} uses an internal
\texttt{tracemin\_lu} solver that returns NaN on disconnected graphs.
We replaced it with \texttt{numpy.linalg.eigvalsh} on the explicit Laplacian,
returning $\lambda_2{=}0$ for disconnected components.
\item \textit{Experiment wall time.}
200~rounds $\times$ 50~robots $\times$ 14~conditions required
${\approx}18$~GPU-hours across multiple SLURM submissions.
\end{enumerate}

\textbf{Future work.}
Three extensions are immediately actionable:
(1)~\emph{push-sum gossip}~\cite{kempe2003} for provably correct weighted
averaging under dropped messages and variable delays;
(2)~\emph{SCAFFOLD control variates}~\cite{karimireddy2020} to correct
client drift from severe non-IID partitions without synchronization;
(3)~physical deployment on differential-drive robots with IEEE~802.11p
ad-hoc communication to validate simulation assumptions.

\begin{thebibliography}{9}
\bibitem{mcmahan2017}
H.~B.\ McMahan et~al., ``Communication-efficient learning of deep networks
from decentralized data,'' \textit{AISTATS}, 2017.
\bibitem{fraboni2022}
Y.~Fraboni et~al., ``A general theory for federated optimization with
asynchronous and heterogeneous clients,'' \textit{TMLR}, 2022.
\bibitem{farhadkhani2022}
S.~Farhadkhani et~al., ``Byzantine machine learning made easy by gradient
pre-aggregation,'' \textit{ICML}, 2022.
\bibitem{lamport1982}
L.~Lamport, R.~Shostak, M.~Pease, ``The Byzantine generals problem,''
\textit{ACM TPLS}, 1982.
\bibitem{kempe2003}
D.~Kempe, A.~Dobra, J.~Gehrke, ``Gossip-based computation of aggregate
information,'' \textit{FOCS}, 2003.
\bibitem{karimireddy2020}
S.~P.\ Karimireddy et~al., ``SCAFFOLD: Stochastic controlled averaging for
federated learning,'' \textit{ICML}, 2020.
\end{thebibliography}

\end{document}
